\documentclass[10pt]{article}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{parskip}

\newcommand{\mc}[1]{\mathcal{#1}}
\renewcommand{\rm}[1]{\mathrm{#1}}
\newcommand{\bb}[1]{\mathbb{#1}}
\newcommand{\N}{\bb{N}}
\newcommand{\Q}{\bb{Q}}
\newcommand{\hj}{\textrm{HJ}}
\newcommand{\vdw}{\textrm{VdW}}
\newcommand{\wnfc}{\textrm{W}\N\textrm{FC}}
\newcommand{\fs}{\textrm{FS}}
\newcommand{\fp}{\textrm{FP}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\U}{\mathcal{U}}
\newcommand{\V}{\mathcal{V}}
\newcommand{\fu}{\forall_\U}
\newcommand{\for}[1]{\forall_{#1}}
\newcommand{\om}{\omega}
\newcommand{\bom}{{(\omega)}}
\newcommand{\blom}{{(<\omega)}}
\newcommand{\lom}{{<\om}}
\newcommand{\ra}{\to}
%\usepackage[lite]{mtpro2}

\title{Ramsey Theory revision questions}
\author{Otto Pyper}
\date{}

\begin{document}
\maketitle

\begin{enumerate}
    \item What is Ramsey Theory all about?
    \item State Ramsey's Theorem.
    \item Prove Ramsey's Theorem.
    \item What type of proof is this called?
    \item Explain how the above generalises to $k$-colourings.
    \item Show that any sequence in a totally ordered set has a monotone subsequence.
    \item State Ramsey's Theorem for $r$-sets.
    \item Prove Ramsey's Theorem for $r$-sets.
    \item Prove that given a collection of points $(i,x_i)$ in the plane, the induced function on some subset of them is either concave or convex.
    \item State Finite Ramsey.
    \item Prove Finite RAmsey.
    \item What type of argument is this called?
    \item Let $c$ be a colouring of $\N^{(2)}$ with an arbitrary set of colours. Then there exists an infinite $M\subset \N$ such that one of which four conditions must hold?
    \item Prove this.
    \item What are the canonical colourings fo the $r$-sets?
    \item Define focused/colour-focused APs.
    \item Define $W(m,k)$.
    \item State Van der Waerden's Theorem.
    \item Prove Van der Waerden's Theorem for $m = 3$.
    \item Prove Van der Waerden's Theorem.
    \item Define the Ackermann/Grzegorcyzk Hierarchy.
    \item Define what it means for a function $f : \N \ra \N$ to be of type $n$.
    \item What is the type of our bound for $W(m,k)$ for fixed $m$?
    \item What is the type of our bound for $W(m) = W(m,2)$?
    \item What did Shelah improve this bound to, and when?
    \item Who offered how much for an $f_3(m)$ bound?
    \item Who found an `almost' type $2$ bound, what was it, and when?
    \item State a lower bound on $W(m)$.
    \item Give two arguments to show that we cannot guarantee infinitely long mono APs.
    \item State the Strengthened Van der Waerden Theorem.
    \item Prove the Strengthened Van der Waerden Theorem.
    \item What kind of bounds do we get here?
    \item What is the case $m = 2$ called?
    \item Prove this directly from Ramsey.
    \item Define the $n$-dimensional cube on alphabet $X$.
    \item Define a combinatorial line.
    \item Define active coordinates.
    \item State the Hales-Jewett Theorem.
    \item Define $\hj(m,k)$.
    \item Show that $\hj \implies \vdw$.
    \item Define what it means for lines to be focused/colour-focused.
    \item Prove the Hales-Jewett Theorem.
    \item Define a $d$-parameter set/$d$-dimensional subspace of $X^n$.
    \item State the Extended Hales-Jewett Theorem.
    \item Prove the Extended Hales-Jewett Theorem.
    \item For $S\subset \N^d$ finite, define a homothetic copy of $S$.
    \item State Gallai's Theorem.
    \item Prove Gallai's Theorem.
    \item Why didnt' we use extended HJ for this, when it might seem so natural?
    \item Define a partition regular matrix $A$.
    \item Express Schur's Theorem as a monochromatic solution to a linear equation.
    \item Express Strengthened VdW similarly.
    \item Define what it means for $A$ to have the columns property.
    \item State Rado's Theorem.
    \item State a handy $(p-1)$-colouring of $\N$, for $p$ prime.
    \item Prove that if $(a_1 \ a_2 \ \dots \ a_n)$ is PR, then $\sum_{i\in I}a_i = 0$ for some $\emptyset\ne I \subset [n]$.
    \item Prove that, for $\lambda \in \Q$, $\wnfc$ there exists mono $x,y,z$ with $x + \lambda y = z$.
    \item Prove Rado's Theorem for a single equation.
    \item State Rado's Boundedness Conjecture.
    \item How many colours suffice for a $1\times 3$? Who proved this, and when?
    \item Show that if $A$ ($m\times n$ rational matrix) is PR, then it has the columns property.
    \item Define an $(m,p,c)$-set
    \item Define a row of an $(m,p,c)$-set.
    \item What is a $(2,p,1)$-set?
    \item What is a $(2,p,3)$-set?
    \item Let $m,p,c\in\N$. Prove that $\wnfc$ there exists a monochromatic $(m,p,c)$-set.
    \item What does the $(m,1,1)$ case mean?
    \item State the Finite Sums Theorem, and all of its other names.
    \item What does this mean for $m = 2$?
    \item Deduce the `Finite Products Theorem'.
    \item What about FS \textit{and} FP?
    \item Show that if matrix $A$ has CP, then there exists $m,p,c$ such that every $(m,p,c)$-set contains a solution of $Ax = 0$.
    \item Prove Rado's Theorem, assuming any earlier results.
    \item State and prove the Consistency Theorem.
    \item Is this obvious?
    \item Can this be proven directly?
    \item Show that $\wnfc$ there exists a colour class containing a solution to every PR system of equations.
    \item State Rado's Conjecture.
    \item Who proved it, and when?
    \item What, roughly speaking, was their method?
    \item State Hindman's Theorem.
    \item Define a filter on $\N$.
    \item Give some examples.
    \item Define an ultrafilter on $\N$.
    \item Give an equivalent characterisation of ultrafilters.
    \item Show that every filter can be extended to an ultrafilter.
    \item An ultrafilter is non-principal iff it extends which filter?
    \item What ingredient do we need in order to actually get an ultrafilter?
    \item Define the topological space $\beta\N$.
    \item How can we view $\N$ as a subset of $\beta\N$?
    \item Show that $\N$ is dense in $\beta\N$.
    \item Prove that $\beta\N$ is Hausdorff.
    \item Prove that $\beta\N$ is compact.
    \item Give an alternative proof of compactness, via Tychonoff.
    \item Why is $\beta\N$ interesting?
    \item Define $\fu x\ p(x)$.
    \item Show that ultrafilter quantifiers play nicely with logical connectives.
    \item Which logical property do ultrafilter quantifiers fail to uphold? Demonstrate this.
    \item Define $\U + \V$, and show it is an ultrafilter.
    \item Show that $+$ on $\beta\N$ is associative.
    \item Show that $+$ is left-continuous.
    \item Show that $+$ is not right-continuous.
    \item Show that $+$ is not commutative.
    \item State the Idempotent Lemma.
    \item Prove the Idempotent Lemma.
    \item Does $\beta\N$ contain any non-trivial finite subgroups?
    \item Prove Hindman's Theorem.
    \item Give a $2$-colouring of $\N^\bom$ for which no $M\subset \N^\bom$ is monochromatic.
    \item Define what it means for $Y\in\N^\bom$ to be Ramsey.
    \item Give a base of open sets for the product topology on $\N^\bom$.
    \item Give two other names for this topology on $\N^\bom$.
    \item Define $(A,M)^\bom$.
    \item Define $M$ accepts $A$.
    \item Define $M$ rejects $A$.
    \item If $M$ accepts/rejects $A$, what can we say about any $L \in M^\bom$?
    \item If $M$ accepts $A$, then $M$ accepts any $A\cup B$ where...?
    \item Need $M$ either accept or reject $A$?
    \item State the Galvin-Prikry Lemma.
    \item Prove the Galvin-Prikry Lemma.
    \item Prove that if $Y\subset \N^\bom$ is open, then it is Ramsey.
    \item Show also that closed sets are Ramsey.
    \item Define the $\ast$-topology on $\N^\bom$, and give two other names for it.
    \item Demonstrate that this really is a base for a topology.
    \item Show that if $Y$ is $\ast$-open, then it is Ramsey.
    \item Hence show also that $\ast$-closed sets are Ramsey.
    \item Define a completely Ramsey set.
    \item Demonstrate that this is strictly stronger than being Ramsey.
    \item Show that $\ast$-open/closed sets are completely Ramsey.
    \item Define a nowhere dense subset of a general topological space.
    \item Show that $\ast$-ND sets are CR.
    \item Define a meagre set.
    \item Show that $\ast$-meagre sets are ND.
    \item Define a Baire set.
    \item Show the Baire sets form a $\sigma$-algebra.
    \item Show that $Y$ is $\ast$-Baire iff it is completely Ramsey.
    \item Why do we bother with meagre sets, since they are just the same as nohwere dense here?
    \item Show that all $\tau$-Borel sets are Ramsey.
    \item Give a neat example of an applicaiton of this result.
\end{enumerate}



\end{document}